Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-12 paper/ms.tex
consistent bibitems

The power spectrum is computed from taking the square of the Fourier transform of the data
time series; hence, the power-spectrum must be non-negative.  In constructing a kernel,
if any of the parameters $a_j$ or $b_j$ are negative, then it is possible for the power
spectrum to go negative, which violates the definition of the power spectrum.  Consequently,
if any of these coefficients is negative, it is necessary to check that the power spectrum
is still non-negative.  Note that a non-negative power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  The positive power spectrum is related to the
positive eigenvalues of the covariance matrix which are necessary for a positive-definite matrix
in limiting cases \citep{messerschmitt2006autocorrelation}.  We find empirically that requiring an everywhere
positive power spectrum results in positive eigenvalues for the covariance matrix, and
so we describe here how to ensure a positive power spectrum using Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

@@

A procedure based upon Sturm's theorem \citep{Dorrie1965} allows one to determine whether there are any
real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.

@@

@@

% A promising technique for addressing both of these goals is the Gaussian Process formalism
% \citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
% correlations is completely specified by a correlation matrix.  In its simplest form, the
% correlation matrix is modeled with a time-independent autocorrelation function (although
% the time-independence can be relaxed), including a white-noise component that may vary
% with each exposure, and the autocorrelation function is described by a functional form that may be
% simply parameterized by a relatively small number of parameters, sometimes referred to as
% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
% describe the autocorrelation function, thus inferring properties of the stellar variabilty
% which may be used to characterize properties of the star, such as the rotation period,
% the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
% The goals of planet detection include measuring the Doppler shift of the star, measuring
% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
% look for dynamical interactions between planets.  To properly account for the impact of
% stellar variability on the planet parameters requires treating correlations in the data, and
% to do so also requires evaluating the correlation matrix.

@@

% One promising solution is to use approximate techiques based upon the fact that correlation
% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
% solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
% solver is complicated and can be challenging to use with general choices of kernel.  Another
% promising approximate technique is to use a uniform resampling of the dataset, and then
% rely on the structure of Toeplitz matrices, which require uniformly space time series,
% to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
% this approach has the drawback of being approximate.  A third approach is to use a kernel
% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
% model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
% of the Lorentzians are required to have a specific relation.  In addition, these
% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

@@

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

@@

% \section{Outline}
% \begin{enumerate}
%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
%\item Assumptions:\\
%  a). Stationary noise (although this can be broken).\\
%  b). Gaussian
%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
%   curves; Solar curves; short-cadence.
%\item Solution(s):\\
%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
%  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
%  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
%\item Solution: Ambikasaran \citep{Ambikasaran:2015} shows how sum of exponential kernels can be solved in order
%   $O(N*p^2)$, where p is number of Lorentzians.
%  - Sum of Lorentzians is a good description of stellar variability:
%     - granulation has expoential (or sums of exponential) noise
%     - asteroseismic obeys sums of Lorentzians
%     - periodic variability due to spots/plages/etc. can also be
%       approximated by Lorentzian(s)
%\item Generalized Ambikasarn for complex coefficients (Hermitian).
% \item Performance:\\
%   a). Show $O(N*p^2)$ scaling\\
%   b). Show comparison to HODLR.\\
%   c). Show comparison to wavelet (ala Carter \& Winn).\\
%   d). Compare with CARMA Kallman filter(?).
% \item Various methods/formulae:\\
%   a). Likelihood computation.\\
%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
%   c). Inferring particular components.\\
%   d). Computing derivatives of GPs (as used in RV methods).\\
%   e). Forecasting/interpolating.\\
%   f). Inferring confidence intervals.\\
%   g). Derivative of likelihood function with respect to kernel parameters.\\
%   h). Non-stationary GP (can the coefficients vary with time?).\\
%   i). Multi-band (each element of extended matrix becomes a matrix).
%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
% \item Generalization to non-stationary noise.
% \item Example application: TYC 3559\\
%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
%   b). Analysis of solar data.
%   c). CO2 Mauna Kea
%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
%   e). TBD Forecasting/interpolating.\\
%   f). TBD Inferring confidence intervals.\\
%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
%   i). TBD Multi-band (each element of extended matrix becomes a matrix).

@@

% \item Possible future applications:\\
%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
%       with Lorentzians, not Gaussians.\\
%   b). Multi-waveband GPs.\\
%   c). RV interpolation.\\
%   d). Non-parametric phase functions.\\
%   e). Better calibration of the flicker log(g) relation.\\
%   f). Measurement of rotational periods - gyrochronology.\\
%   g). Doppler shifts in oscillating EBs.

@@

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,Michel:2009}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
\begin{equation}
\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be
expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
deviation of 0.1\%.
\end{itemize}

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.    Within astrophysics, correlated noise (due to the environment,
detector, or modeling uncertainty) may be present in
gravitational wave time series, and so Gaussian processes may be a way to address
this problem \citep{Moore2016}.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.

Outside of astronomy, this technique may have application to seismology \citep{Robinson1967},

@@

\facility{Kepler}
\software{%
    % \project{ceres} \citep{Agarwal:2016},
    % \project{corner.py} \citep{Foreman-Mackey:2016},
    % \project{emcee} \citep{Foreman-Mackey:2013},
    % \project{george} \citep{Ambikasaran2016},
	% \project{matplotlib} \citep{Hunter:2007},
	% \project{numpy} \citep{Van-Der-Walt:2011},
	% \project{scipy} \citep{Jones:2001}.
}

@@

As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with the $d_j=0$
terms first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{u_{2,j},g_{2,j};j=1,...,J_0\}, \{u_{2,j},v_{2,j},g_{2,j},h_{2,j};j=J_0+1,...,J\},
x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [y_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, y_2/2, ..., y_N/2]$.
The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
with a banded algorithm.

>>>>

The power spectrum is computed from taking the square of the Fourier transform
of the data time series; hence, the power-spectrum must be non-negative.  In
constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
then it is possible for the power spectrum to go negative, which violates the
definition of the power spectrum.
Consequently, if any of these coefficients is negative, it is necessary to
check that the power spectrum is still non-negative.  Note that a non-negative
power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  The positive power spectrum is
related to the positive eigenvalues of the covariance matrix which are
necessary for a positive-definite matrix in limiting cases
\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
positive power spectrum results in positive eigenvalues for the covariance
matrix, and so we describe here how to ensure a positive power spectrum using
Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

@@

A procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.

@@

KIC~11615890 \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz}
\end{eqnarray}

@@

% A promising technique for addressing both of these goals is the Gaussian Process formalism
% \citep{Rasmussen2006,Gibson:2012}.  The basis of this method is to assume that stellar variability
% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
% correlations is completely specified by a correlation matrix.  In its simplest form, the
% correlation matrix is modeled with a time-independent autocorrelation function (although
% the time-independence can be relaxed), including a white-noise component that may vary
% with each exposure, and the autocorrelation function is described by a functional form that may be
% simply parameterized by a relatively small number of parameters, sometimes referred to as
% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
% describe the autocorrelation function, thus inferring properties of the stellar variabilty
% which may be used to characterize properties of the star, such as the rotation period,
% the asteroseismic frequencies, or granulation noise known as `flicker'
% \citep{Aerts2010,Noyes:1984,Bastien:2013}.
% The goals of planet detection include measuring the Doppler shift of the star, measuring
% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
% look for dynamical interactions between planets.  To properly account for the impact of
% stellar variability on the planet parameters requires treating correlations in the data, and
% to do so also requires evaluating the correlation matrix.

@@

% One promising solution is to use approximate techiques based upon the fact that correlation
% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
% solver is only approximate \citep{Ambikasaran:2013,Ambikasaran:2016}.  In addition, the
% solver is complicated and can be challenging to use with general choices of kernel.  Another
% promising approximate technique is to use a uniform resampling of the dataset, and then
% rely on the structure of Toeplitz matrices, which require uniformly space time series,
% to obtain approximate solutions of the likelihood function \citep{Wilson:2015},
% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
% this approach has the drawback of being approximate.  A third approach is to use a kernel
% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
% model \citep[aka \texttt{CARMA};][]{Kelly:2014}.  This approach can model a stationary power-spectrum that
% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
% of the Lorentzians are required to have a specific relation.  In addition, these
% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

@@

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

@@

% \section{Outline}
% \begin{enumerate}
%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
%\item Assumptions:\\
%  a). Stationary noise (although this can be broken).\\
%  b). Gaussian
%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
%   curves; Solar curves; short-cadence.
%\item Solution(s):\\
%  a). HODLR \citep{Ambikasaran:2013,Ambikasaran:2016}\\
%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly:2014}\\
%  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
%  d). KISS-GP \citep{Wilson:2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
%\item Solution: Ambikasaran \citep{Ambikasaran:2015} shows how sum of exponential kernels can be solved in order
%   $O(N*p^2)$, where p is number of Lorentzians.
%  - Sum of Lorentzians is a good description of stellar variability:
%     - granulation has expoential (or sums of exponential) noise
%     - asteroseismic obeys sums of Lorentzians
%     - periodic variability due to spots/plages/etc. can also be
%       approximated by Lorentzian(s)
%\item Generalized Ambikasarn for complex coefficients (Hermitian).
% \item Performance:\\
%   a). Show $O(N*p^2)$ scaling\\
%   b). Show comparison to HODLR.\\
%   c). Show comparison to wavelet (ala Carter \& Winn).\\
%   d). Compare with CARMA Kallman filter(?).
% \item Various methods/formulae:\\
%   a). Likelihood computation.\\
%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
%   c). Inferring particular components.\\
%   d). Computing derivatives of GPs (as used in RV methods).\\
%   e). Forecasting/interpolating.\\
%   f). Inferring confidence intervals.\\
%   g). Derivative of likelihood function with respect to kernel parameters.\\
%   h). Non-stationary GP (can the coefficients vary with time?).\\
%   i). Multi-band (each element of extended matrix becomes a matrix).
%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
% \item Generalization to non-stationary noise.
% \item Example application: TYC 3559\\
%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
%   b). Analysis of solar data.
%   c). CO2 Mauna Kea
%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
%   e). TBD Forecasting/interpolating.\\
%   f). TBD Inferring confidence intervals.\\
%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
%   i). TBD Multi-band (each element of extended matrix becomes a matrix).

@@

% \item Possible future applications:\\
%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{Wilson:2013}, but
%       with Lorentzians, not Gaussians.\\
%   b). Multi-waveband GPs.\\
%   c). RV interpolation.\\
%   d). Non-parametric phase functions.\\
%   e). Better calibration of the flicker log(g) relation.\\
%   f). Measurement of rotational periods - gyrochronology.\\
%   g). Doppler shifts in oscillating EBs.

@@

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
The power spectrum of a single exponential scales as $f^{-2}$ at high
frequency, while the Mat\'ern kernel scales as $f^{-4}$ which more
closely matches the power spectrum due to activity and granulation in
the Sun \citep{Aigrain:2004,Michel:2009}.
\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
\begin{equation}
k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
\end{equation}
where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
\begin{equation}
\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
\end{equation}
This kernel may be approximated well by the sum of four exponential kernels
\begin{eqnarray}
k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
 &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
\end{eqnarray}
which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
which has been used to model quasi-periodic stellar variability due to star spots.   It is
given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be
expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:
\begin{equation}
E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
\end{equation}
with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
deviation of 0.1\%.
\end{itemize}

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.    Within astrophysics, correlated noise (due to the environment,
detector, or modeling uncertainty) may be present in
gravitational wave time series, and so Gaussian processes may be a way to address
this problem \citep{Moore:2016}.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011}, and modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}.

Outside of astronomy, this technique may have application to seismology
\citep{Robinson:1967},

@@

\facility{Kepler}
\software{%
    % \project{ceres} \citep{Agarwal:2016},
    % \project{corner.py} \citep{Foreman-Mackey:2016},
    % \project{emcee} \citep{Foreman-Mackey:2013},
    % \project{george} \citep{Ambikasaran:2016},
	% \project{matplotlib} \citep{Hunter:2007},
	% \project{numpy} \citep{Van-Der-Walt:2011},
	% \project{scipy} \citep{Jones:2001}.
}

@@

As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
solved with a banded solver, as in \citet{Press:1992}.  The determinant of this extended matrix
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with the $d_j=0$
terms first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{u_{2,j},g_{2,j};j=1,...,J_0\}, \{u_{2,j},v_{2,j},g_{2,j},h_{2,j};j=J_0+1,...,J\},
x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [y_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, y_2/2, ..., y_N/2]$.
The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
with a banded algorithm.
